---
output:
  pdf_document: default
  html_document: default
---

## Setup and load libraries

```{r libraries}
library(haven)
library(rstanarm)
library(performance)
library(lme4)
library(glmnet)
library(moments)
library(tidyverse)
```

## Loading data

```{r}
load("/T23/bsaengm8.rdata")   # → BSAENGM8
load("/T23/bsgengm8.rdata")   # → BSGENGM8
load("/T23/bcgengm8.rdata")   # → BCGENGM8
```

```{r}
# Strip SPSS labels
BSAENGM8 <- zap_labels(BSAENGM8)
BSGENGM8 <- zap_labels(BSGENGM8)
BCGENGM8 <- zap_labels(BCGENGM8)
```

---

## Data Cleaning & Variable Construction

### Helper Function

```{r}
# Recode 9/99/999 to NA
clean_val <- function(x, missing = c(9, 99, 999, 9999)) {
  x <- as.numeric(x)
  x[x %in% missing] <- NA
  x
}
```

### We average over the 5 Science Score values 

```{r}
# Average across 5 plausible values
BSAENGM8 <- BSAENGM8 %>%
  mutate(PV1SCIE = rowMeans(select(., BSSSCI01:BSSSCI05), na.rm = TRUE))
```

### encoding: Student-Level Variables

```{r}
BSGENGM8 <- BSGENGM8 %>%
  mutate(
    GENDER = factor(clean_val(ITSEX, 9),
                    levels = c(1, 2),
                    labels = c("Girl", "Boy")),
    ESCS = factor(clean_val(BSDGHER, 9),
                  levels = c(1, 2, 3), ordered = TRUE,
                  labels = c("Many Resources", "Some Resources", "Few Resources")),
    BOOKS_HOME = factor(clean_val(BSBG04, 9),
                        levels = 1:5, ordered = TRUE,
                        labels = c("0-10", "11-25", "26-100", "101-200", "200+")),
    Internet_or_room = factor(clean_val(BSDG05S, 9),
                              levels = c(1, 2, 3), ordered = TRUE,
                              labels = c("None", "Either", "Both")),
    PAR_EDU = factor(clean_val(BSDGEDUP, c(6, 9)),
                     levels = 1:5, ordered = TRUE,
                     labels = c("University+", "Post-secondary", "Upper secondary",
                                "Lower secondary", "Some primary")),
    BELONGING = factor(clean_val(BSDGSSB, 9),
                       levels = c(1, 2, 3), ordered = TRUE,
                       labels = c("High", "Some", "Little"))
  )
```

### School-Level Variables

```{r}
BCGENGM8 <- BCGENGM8 %>%
  mutate(
    SCH_DISADV = factor(clean_val(BCBG03A, 9),
                        levels = 1:4, ordered = TRUE,
                        labels = c("0-10%", "11-25%", "26-50%", "50%+")),
    SCH_SES = factor(clean_val(BCDGSBC, 9),
                     levels = c(1, 2, 3),
                     labels = c("More Affluent", "Mixed", "More Disadvantaged")),
    ACAD_EMPHASIS = factor(clean_val(BCDGEAS, 9),
                           levels = c(1, 2, 3), ordered = TRUE,
                           labels = c("Very High", "High", "Medium"))
  )
```


## 4. Merging & Final Dataset

```{r}
df <- BSAENGM8 %>%
  select(IDSCHOOL, IDSTUD, PV1SCIE) %>%
  left_join(
    BSGENGM8 %>% select(IDSCHOOL, IDSTUD, GENDER, ESCS, BOOKS_HOME,
                        Internet_or_room, PAR_EDU, BELONGING),
    by = c("IDSCHOOL", "IDSTUD")
  ) %>%
  left_join(
    BCGENGM8 %>% select(IDSCHOOL, SCH_SES, SCH_DISADV, ACAD_EMPHASIS),
    by = "IDSCHOOL"
  )
```


```{r}
# Drop students in schools with no school questionnaire response
df_clean <- df %>%
  filter(!is.na(SCH_SES)) %>%
  droplevels()
```


### Missingness Check

```{r}
df_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(pct = round(n_missing / nrow(df_clean) * 100, 1)) %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))
```
Missigness to high for PAR_EDU - this variable will not be used as a predictor. 


```{r}
# Number of students
nrow(df_clean)

# Number of schools
length(unique(df_clean$IDSCHOOL))
```


```{r}
head(df_clean)
```

```{r}
summary(df_clean)
```
Proceed with our cleaned dataset 


## 5. Exploratory Data Analysis
### Visualisations

```{r}
# Summary statistics to support Gaussian assumption
cat("Mean:", round(mean(df_clean$PV1SCIE), 1), "\n")
cat("Median:", round(median(df_clean$PV1SCIE), 1), "\n")
cat("Skewness:", round(moments::skewness(df_clean$PV1SCIE), 3), "\n")

# Histogram with normal curve overlay
ggplot(df_clean, aes(x = PV1SCIE)) +
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 30, fill = "#378ADD", alpha = 0.7) +
  stat_function(fun = dnorm, 
                args = list(mean = mean(df_clean$PV1SCIE), 
                            sd = sd(df_clean$PV1SCIE)),
                colour = "#D85A30", linewidth = 1) +
  labs(title = "Distribution of science scores with normal curve",
       x = "Science Score", y = "Density") +
  theme_minimal(base_size = 12)
```

```{r}
table(df_clean$GENDER)
prop.table(table(df_clean$GENDER))
```

```{r}
p2 <- ggplot(df_clean, aes(x = GENDER, y = PV1SCIE, fill = GENDER)) +
  geom_boxplot(outlier.size = 0.8, outlier.alpha = 0.4, width = 0.5) +
  scale_fill_manual(values = c("Girl" = "#378ADD", "Boy" = "#1D9E75")) +
  labs(
    title = "Science Score by Gender",
    x     = NULL,
    y     = "Science Score"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

p2
```

```{r}
df_clean %>%
  group_by(GENDER) %>%
  summarise(
    median_score = round(median(PV1SCIE, na.rm = TRUE), 1),
    Q1 = round(quantile(PV1SCIE, 0.25, na.rm = TRUE), 1),
    Q3 = round(quantile(PV1SCIE, 0.75, na.rm = TRUE), 1)
  )
```

```{r}
df_clean %>%
  summarise(
    mean_score = round(mean(PV1SCIE, na.rm = TRUE), 1)
  )
```


```{r}
p5 <- ggplot(df_clean %>% filter(!is.na(ESCS)),
             aes(x = ESCS, y = PV1SCIE, fill = GENDER)) +
  geom_boxplot(outlier.size = 0.8, outlier.alpha = 0.4, width = 0.6,
               position = position_dodge(0.75)) +
  scale_fill_manual(values = c("Girl" = "#378ADD", "Boy" = "#1D9E75")) +
  scale_x_discrete(limits = c("Many Resources", "Some Resources", "Few Resources")) +
  labs(
    title = "Science Score by Gender and Home Resources",
    x     = "Home Educational Resources",
    y     = "Science Score",
    fill  = "Gender"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

p5
```
#more plots but weren't used in the report 

```{r}

# Plot 1: Books at home
p1 <- ggplot(df_clean %>% filter(!is.na(BOOKS_HOME), !is.na(GENDER)),
             aes(x = BOOKS_HOME, fill = GENDER)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("Girl" = "#378ADD", "Boy" = "#1D9E75")) +
  labs(title = "Books at home", x = NULL, y = "Count", fill = "Gender") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none")


# Plot 3: Parental education
p3 <- ggplot(df_clean %>% filter(!is.na(PAR_EDU), !is.na(GENDER)),
             aes(x = PAR_EDU, fill = GENDER)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("Girl" = "#378ADD", "Boy" = "#1D9E75")) +
  labs(title = "Parental education", x = NULL, y = "Count", fill = "Gender") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none")

# Plot 4: School disadvantage
p4 <- ggplot(df_clean %>% filter(!is.na(SCH_DISADV), !is.na(GENDER)),
             aes(x = SCH_DISADV, fill = GENDER)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("Girl" = "#378ADD", "Boy" = "#1D9E75")) +
  labs(title = "School % disadvantaged", x = NULL, y = "Count", fill = "Gender") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p1
p4
p3

```



```{r}
# Quick plot
ggplot(df_clean %>% filter(!is.na(BELONGING), !is.na(GENDER)),
       aes(x = BELONGING, fill = GENDER)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("Girl" = "#378ADD", "Boy" = "#1D9E75")) +
  labs(title = "Sense of school belonging by gender",
       x = "Belonging", y = "Count", fill = "Gender") +
  theme_minimal(base_size = 11)
```
```{r}
df_clean %>%
  filter(!is.na(BELONGING)) %>%
  group_by(GENDER) %>%
  summarise(
    pct_little = round(mean(BELONGING == "Little") * 100, 1),
    pct_some   = round(mean(BELONGING == "Some") * 100, 1),
    pct_high   = round(mean(BELONGING == "High") * 100, 1),
    n = n()
  )
```


## 6. Modelling

### 6.1 Prepare Numeric Variables

```{r}
df_model <- df_clean %>%
  mutate(
    # GENDER: Girl = 0, Boy = 1
    GENDER_num     = as.integer(GENDER == "Boy"),

    # ESCS: reverse so higher = more resources
    ESCS_num       = 4L - as.integer(ESCS),

    # BOOKS_HOME: already low to high
    BOOKS_num      = as.integer(BOOKS_HOME),

    # Internet_or_room: already low to high
    INTERNET_num   = as.integer(Internet_or_room),

    # PAR_EDU: reverse so higher = more education
    PAR_EDU_num    = ifelse(PAR_EDU == "Not applicable", NA_real_,
                            6L - as.integer(PAR_EDU) + 1L),

    # BELONGING: reverse so higher = more belonging
    BELONGING_num  = 4L - as.integer(BELONGING),

    # SCH_DISADV: already low to high
    SCH_DISADV_num = as.integer(SCH_DISADV),

    # ACAD_EMPHASIS: reverse so higher = more emphasis
    ACAD_num       = 4L - as.integer(ACAD_EMPHASIS),

    # SCH_SES: dummy coded, reference = More Affluent
    SCH_SES_mixed  = as.integer(SCH_SES == "Mixed"),
    SCH_SES_disadv = as.integer(SCH_SES == "More Disadvantaged")
  )
```

### 6.2 Missing-Indicator Imputation

```{r}
df_model <- df_model %>%
  mutate(
    BELONGING_missing = as.integer(is.na(BELONGING_num)),
    INTERNET_missing  = as.integer(is.na(INTERNET_num)),
    ESCS_missing      = as.integer(is.na(ESCS_num)),
    BOOKS_missing     = as.integer(is.na(BOOKS_num)),

    # Replace NAs with 0 so rows are retained
    BELONGING_num = replace_na(BELONGING_num, 0),
    INTERNET_num  = replace_na(INTERNET_num, 0),
    ESCS_num      = replace_na(ESCS_num, 0),
    BOOKS_num     = replace_na(BOOKS_num, 0)
  )
```

Standardise the PV1SCIE to obtain simpler parameters for the prior 
```{r}
df_model <- df_model %>%
  mutate(PV1SCIE_std = as.numeric(scale(PV1SCIE)))
```


### 6.4 Frequentist Models

#### Model 1 — Full (with all missing flags)

```{r}
model_freq_1 <- lm(
  PV1SCIE_std ~ GENDER_num + ESCS_num + BOOKS_num + INTERNET_num +
    BELONGING_num + SCH_DISADV_num + ACAD_num +
    SCH_SES_mixed + SCH_SES_disadv +
    BELONGING_missing + INTERNET_missing +
    ESCS_missing + BOOKS_missing,
  data = df_model
)

summary(model_freq_1)
```

#### Model 2 — Final (backward selection; non-significant terms dropped)

```{r}
model_freq_final <- lm(
  PV1SCIE_std ~ GENDER_num + ESCS_num + BOOKS_num + BELONGING_num +
    SCH_DISADV_num + ACAD_num + SCH_SES_mixed +
    BELONGING_missing + BOOKS_missing,
  data = df_model
)

summary(model_freq_final)
```

```{r}
icc_manual <- function(grouping_var) {
  formula <- as.formula(paste("PV1SCIE_std ~ 1 + (1 |", grouping_var, ")"))
  m <- lmer(formula, data = df_model)
  vc <- as.data.frame(VarCorr(m))
  between <- vc$vcov[1]
  total   <- sum(vc$vcov)
  cat(grouping_var, "ICC =", round(between / total, 4), "\n")
}

icc_manual("SCH_DISADV")
icc_manual("SCH_SES")
icc_manual("ACAD_EMPHASIS")
```

### 6.5 Bayesian Model

```{r}
multilevel_intercept <- stan_lmer(
  PV1SCIE_std ~ GENDER_num + ESCS_num + BOOKS_num + BELONGING_num +
                SCH_DISADV_num + ACAD_num +
                BELONGING_missing + BOOKS_missing +
                (1 | SCH_DISADV),
  data            = df_model,
  prior           = normal(location = 0, scale = 2.5),
  prior_intercept = normal(location = 0, scale = 5),
  chains          = 4,
  iter            = 2000,
  seed            = 1234,
  refresh         = 0
)

summary(multilevel_intercept)
posterior_interval(multilevel_intercept, prob = 0.95)
summary(bayes_R2(multilevel_intercept))
```



#testing interactions 

```{r}

# BOOKS x SCH_SES
m2_books_ses <- lm(PV1SCIE_std ~ BOOKS_num * SCH_SES + GENDER_num + 
                   ESCS_num + BELONGING_num + SCH_DISADV_num + ACAD_num +
                   BELONGING_missing + BOOKS_missing,
                   data = df_model)
anova(model_freq_final , m2_books_ses)

# ESCS x SCH_SES
m2_escs_ses <- lm(PV1SCIE_std ~ ESCS_num * SCH_SES + GENDER_num + 
                  BOOKS_num + BELONGING_num + SCH_DISADV_num + ACAD_num +
                  BELONGING_missing + BOOKS_missing,
                  data = df_model)
anova(model_freq_final, m2_escs_ses)

# GENDER x SCH_SES
m2_gender_ses <- lm(PV1SCIE_std ~ GENDER_num * SCH_SES + ESCS_num + 
                    BOOKS_num + BELONGING_num + SCH_DISADV_num + ACAD_num +
                    BELONGING_missing + BOOKS_missing,
                    data = df_model)
anova(model_freq_final , m2_gender_ses)

# BELONGING x SCH_SES
m2_belonging_ses <- lm(PV1SCIE_std ~ BELONGING_num * SCH_SES + GENDER_num + 
                       ESCS_num + BOOKS_num + SCH_DISADV_num + ACAD_num +
                       BELONGING_missing + BOOKS_missing,
                       data = df_model)
anova(model_freq_final, m2_belonging_ses)


# BOOKS x SCH_DISADV
m2_books_disadv <- lm(PV1SCIE_std ~ BOOKS_num * SCH_DISADV + GENDER_num + 
                      ESCS_num + BELONGING_num + ACAD_num + SCH_SES_mixed +
                      SCH_SES_disadv + BELONGING_missing + BOOKS_missing,
                      data = df_model)
anova(model_freq_final, m2_books_disadv)

# BOOKS x ACAD_EMPHASIS
m2_books_acad <- lm(PV1SCIE_std ~ BOOKS_num * ACAD_EMPHASIS + GENDER_num + 
                    ESCS_num + BELONGING_num + SCH_DISADV_num + SCH_SES_mixed +
                    SCH_SES_disadv + BELONGING_missing + BOOKS_missing,
                    data = df_model)
anova(model_freq_final, m2_books_acad)

# ESCS x SCH_DISADV
m2_escs_disadv <- lm(PV1SCIE_std ~ ESCS_num * SCH_DISADV + GENDER_num + 
                     BOOKS_num + BELONGING_num + ACAD_num + SCH_SES_mixed +
                     SCH_SES_disadv + BELONGING_missing + BOOKS_missing,
                     data = df_model)
anova(model_freq_final, m2_escs_disadv)

# ESCS x ACAD_EMPHASIS
m2_escs_acad <- lm(PV1SCIE_std ~ ESCS_num * ACAD_EMPHASIS + GENDER_num + 
                   BOOKS_num + BELONGING_num + SCH_DISADV_num + SCH_SES_mixed +
                   SCH_SES_disadv + BELONGING_missing + BOOKS_missing,
                   data = df_model)
anova(model_freq_final, m2_escs_acad)

# GENDER x SCH_DISADV
m2_gender_disadv <- lm(PV1SCIE_std ~ GENDER_num * SCH_DISADV + ESCS_num + 
                       BOOKS_num + BELONGING_num + ACAD_num + SCH_SES_mixed +
                       SCH_SES_disadv + BELONGING_missing + BOOKS_missing,
                       data = df_model)
anova(model_freq_final, m2_gender_disadv)

# GENDER x ACAD_EMPHASIS
m2_gender_acad <- lm(PV1SCIE_std ~ GENDER_num * ACAD_EMPHASIS + ESCS_num + 
                     BOOKS_num + BELONGING_num + SCH_DISADV_num + SCH_SES_mixed +
                     SCH_SES_disadv + BELONGING_missing + BOOKS_missing,
                     data = df_model)
anova(model_freq_final, m2_gender_acad)

# BELONGING x SCH_DISADV
m2_belonging_disadv <- lm(PV1SCIE_std ~ BELONGING_num * SCH_DISADV + GENDER_num + 
                          ESCS_num + BOOKS_num + ACAD_num + SCH_SES_mixed +
                          SCH_SES_disadv + BELONGING_missing + BOOKS_missing,
                          data = df_model)
anova(model_freq_final, m2_belonging_disadv)

# BELONGING x ACAD_EMPHASIS
m2_belonging_acad <- lm(PV1SCIE_std ~ BELONGING_num * ACAD_EMPHASIS + GENDER_num + 
                        ESCS_num + BOOKS_num + SCH_DISADV_num + SCH_SES_mixed +
                        SCH_SES_disadv + BELONGING_missing + BOOKS_missing,
                        data = df_model)
anova(model_freq_final, m2_belonging_acad)

```


we use female & SCH_DISADV
## random slopes

```{r}
multilevel_slopes_v2 <- stan_lmer(
  PV1SCIE_std ~ GENDER_num + ESCS_num + BOOKS_num + BELONGING_num +
                SCH_DISADV_num + ACAD_num +
                BELONGING_missing + BOOKS_missing +
                (1 + GENDER_num | SCH_DISADV),
  data            = df_model,
  prior           = normal(location = 0, scale = 2.5),
  prior_intercept = normal(location = 0, scale = 5),
  chains          = 4,
  iter            = 2000,
  seed            = 1234,
  refresh         = 0
)
```


```{r}
summary(multilevel_slopes_v2)
posterior_interval(multilevel_slopes_v2, prob = 0.95)
summary(bayes_R2(multilevel_slopes_v2))
```



```{r}
set.seed(1234)
n          <- nrow(df_model)
train_idx  <- sample(1:n, size = 0.8 * n)
train_data <- df_model[train_idx, ]
test_data  <- df_model[-train_idx, ]

vars_used <- c("GENDER_num", "ESCS_num", "BOOKS_num", "BELONGING_num",
               "SCH_DISADV_num", "ACAD_num",
               "BELONGING_missing", "BOOKS_missing")

test_eval  <- test_data  %>% drop_na(all_of(c("PV1SCIE_std", vars_used)))
train_eval <- train_data %>% drop_na(all_of(c("PV1SCIE_std", vars_used)))

# Frequentist
freq_train <- lm(
  PV1SCIE_std ~ GENDER_num + ESCS_num + BOOKS_num + BELONGING_num +
                SCH_DISADV_num + ACAD_num +
                BELONGING_missing + BOOKS_missing,
  data = train_eval
)

# Bayesian
bayes_train <- stan_glm(
  PV1SCIE_std ~ GENDER_num + ESCS_num + BOOKS_num + BELONGING_num +
                SCH_DISADV_num + ACAD_num +
                BELONGING_missing + BOOKS_missing,
  data = train_eval,
  prior = normal(0, 2.5),
  prior_intercept = normal(0, 5),
  chains = 4, iter = 2000, seed = 1234, refresh = 0
)

# Multilevel intercept
multi_intercept_train <- stan_lmer(
  PV1SCIE_std ~ GENDER_num + ESCS_num + BOOKS_num + BELONGING_num +
                SCH_DISADV_num + ACAD_num +
                BELONGING_missing + BOOKS_missing +
                (1 | SCH_DISADV),
  data = train_eval,
  prior = normal(0, 2.5),
  prior_intercept = normal(0, 5),
  chains = 4, iter = 2000, seed = 1234, refresh = 0
)

# Multilevel slopes
multi_slopes_train <- stan_lmer(
  PV1SCIE_std ~ GENDER_num + ESCS_num + BOOKS_num + BELONGING_num +
                SCH_DISADV_num + ACAD_num +
                BELONGING_missing + BOOKS_missing +
                (1 + GENDER_num | SCH_DISADV),
  data = train_eval,
  prior = normal(0, 2.5),
  prior_intercept = normal(0, 5),
  chains = 4, iter = 2000, seed = 1234, refresh = 0
)

# Lasso and Ridge

X_train <- model.matrix(PV1SCIE_std ~ .,
                        data = train_eval[, c("PV1SCIE_std", vars_used)])[, -1]
y_train <- train_eval$PV1SCIE_std
X_test  <- model.matrix(PV1SCIE_std ~ .,
                        data = test_eval[, c("PV1SCIE_std", vars_used)])[, -1]
y_test  <- test_eval$PV1SCIE_std

lasso_model <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 10)
ridge_model <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 10)

# Predictions
freq_pred         <- predict(freq_train, newdata = test_eval)
bayes_pred        <- colMeans(posterior_epred(bayes_train, newdata = test_eval))
multi_int_pred    <- colMeans(posterior_epred(multi_intercept_train, newdata = test_eval))
multi_slopes_pred <- colMeans(posterior_epred(multi_slopes_train, newdata = test_eval))
lasso_pred        <- as.numeric(predict(lasso_model, newx = X_test, s = "lambda.min"))
ridge_pred        <- as.numeric(predict(ridge_model, newx = X_test, s = "lambda.min"))

# RMSE
rmse_fn <- function(actual, predicted) sqrt(mean((actual - predicted)^2, na.rm = TRUE))

data.frame(
  Model = c("Frequentist", "Bayesian", "Multilevel intercept",
            "Multilevel slopes", "Lasso", "Ridge"),
  RMSE  = round(c(
    rmse_fn(y_test, freq_pred),
    rmse_fn(y_test, bayes_pred),
    rmse_fn(y_test, multi_int_pred),
    rmse_fn(y_test, multi_slopes_pred),
    rmse_fn(y_test, lasso_pred),
    rmse_fn(y_test, ridge_pred)
  ), 4)
)

```



```{r}
coef(ridge_model, s = "lambda.min")
```

#plot for observed vs predicted 
```{r}


# --- Observed vs predicted (multilevel slopes model) ---
test_eval$predicted <- multi_slopes_pred

obs_pred_plot <- ggplot(test_eval, aes(x = predicted, y = PV1SCIE_std)) +
  geom_point(alpha = 0.3, size = 0.8, colour = "#378ADD") +
  geom_abline(slope = 1, intercept = 0,
              colour = "#D85A30", linewidth = 0.8) +
  labs(title = "Observed vs predicted science scores",
       x     = "Predicted (standardised)",
       y     = "Observed (standardised)") +
  theme_minimal(base_size = 12)

obs_pred_plot


```

```{r}
coef(lasso_model, s = "lambda.min")
```


